Reducing Acceptance Marks in Emerson-Lei Automata by QBF Solving

### 2023-06-04 paper/main.tex
Plots
heatmap tables polished

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G}) from the Spot library~\cite{duret.16.atva2}
and \texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA
by \texttt{ltl3tela} with option \texttt{-D1} and by two
state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva}, namely \texttt{ltl2dela} (known as
Delag)~\cite{muller.17.gandalf} and \texttt{ltl2dgra} (known as
Rabinizer~4)~\cite{kretinsky.18.cav}. Some translators failed on some
formulas: it usually reached a timeout of 60 seconds or it produced an
automaton that cannot be parsed by the Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

  \caption{The cumulative numbers of acceptance marks before and after
    reduction for various reduction levels and automata sets given by
    the translator and formula set. The column \emph{reduction} shows
    the cumulative percentage of saved acceptance marks and
    \emph{time} reports the cumulative reduction time. The column
    \textit{solver timeout} indicates the number of instances where
    the last call to the QBF solver timed out.  }

@@

\fixme{Upravit zbytek textu a popis Table~3.}
Table~\ref{tab:heatmap} shows the number of automata from all
four sets that have a given number of original acceptance marks and a
given number of acceptance marks after multi-level reduction. The
table indicates that in many cases only 1 or 2~marks can be
saved. However, the achieved reduction is substantial for some
automata with a higher number of original acceptance marks. For
example, in 24 cases, we have reduced 7 or more
acceptance marks to only 4.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.

\begin{table}[t]
\caption{The effect of multi-level reduction.
A cell on coordinates $(x,y)$ contains the number of automata that have been
reduced from $x$ to $y$ acceptance marks. If the cell contains a sum of two
numbers, the latter represents the number of automata where the attempt to reduce
another mark has been unsuccessful due to a QBF solver timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap2.tex}
\end{table}

@@

@@

>>>>

To evaluate our reduction technique, we applied \telatko to automata
produced by the following process. We started with two sets of LTL formulas.
\begin{itemize}
\item One set contains all LTL formulas from \emph{literature} that
  are provided by the tool \texttt{genltl} of the Spot
  library~\cite{duret.16.atva2}~2.10.4. For parametrized formula
  patterns, we consider instances for all combinations of parameter
  values from 1 to 4.
\item The second set consists of 400 random LTL formulas with 4 atomic
  propositions. These formulas were generated by the tool
  \texttt{randltl} of the Spot library.
\end{itemize}
On both these sets, we applied the tool \texttt{ltlfilt} of the Spot
library to simplify the formulas and remove duplicates and formulas
equivalent to $\true$ and $\false$. After these steps, we had 348
LTL formulas from literature and 335 random formulas.
%
Formulas from both sets have been translated to nondeterministic TELA
by two state-of-the-art translators, namely \texttt{ltl2tgba} (used
with option \texttt{-G} to get generic TELA) from the Spot
library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}, and to deterministic TELA by
\texttt{ltl3tela} with option \texttt{-D1} and by two state-of-the-art
translators from the Owl library~\cite{kretinsky.18.atva}, namely
\texttt{ltl2dela} (known as Delag)~\cite{muller.17.gandalf} and
\texttt{ltl2dgra} (known as Rabinizer~4)~\cite{kretinsky.18.cav}. Some
translators failed on some formulas: it usually reached a timeout of
60 seconds or it produced an automaton that cannot be parsed by the
Spot library.
% produced by the individual tools contains less than 348 automata as not
% all of the formulas are successfully translated due to a timeout of 60
% seconds.
Further, we have removed automata with 0 or 1 acceptance mark as there
is a little point in reducing these. Table \ref{tab:ltltranslators}
shows the exact versions of the translators. For each translator and
each set of formulas, the table also provides the number of fails, the
number of produced automata with less than two marks, and the number
of automata with at least two marks. The numbers of automata with at
least two marks are typeset in bold as these automata are actually
used for the experimental evaluation of our reduction technique.

@@

  \caption{For each automata set identified by the translator and the
    set of formulas, the table provides the cumulative number of
    acceptance marks before any reduction (in the box), after
    reduction of individual levels and after multi-level reduction
    (column \emph{marks}). The column \emph{reduction} shows the
    percentage of saved acceptance marks and \emph{time} reports the
    cumulative reduction time in seconds. The column \textit{solver
      timeout} indicates the number of instances where the last call
    to the QBF solver timed out.  }

@@

Table~\ref{tab:heatmap} shows the number of automata from all five
sets constructed from formulas from literature that have a given
number of original acceptance marks and a given number of acceptance
marks after multi-level reduction. The table indicates that in many
cases only 1 or 2~marks can be saved. However, the achieved reduction
is substantial for some automata with a higher number of original
acceptance marks. For example, in 26 cases, we have reduced 7 or more
acceptance marks to only 4 or less.
% Moreover, there were many timeouts of the last QBF query for manof these automata.
% Let us note that some that some reductions are the results are promising, e.g. in 17 cases, we have successfully
% reduced 8 or more acceptance marks to only 4.
Table~\ref{tab:heatmap-random} shows the same information for all
considered automata constructed from random formulas.

\begin{table}[t]
  \caption{The effect of multi-level reduction on automata constructed
    from formulas from literature.  A cell on coordinates $(x,y)$
    contains the number of automata that have been reduced from $x$ to
    $y$ acceptance marks. If the cell contains a sum of two numbers,
    the latter represents the number of automata where the attempt to
    reduce another mark has been unsuccessful due to a QBF solver
    timeout.}
\label{tab:heatmap}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap.tex}
\end{table}

\begin{table}[t]
\caption{The effect of multi-level reduction on automata constructed
    from random formulas. The meaning of each cell is the same as in Table~\ref{tab:heatmap}.}
\label{tab:heatmap-random}
\setlength{\tabcolsep}{2.5pt}
\centering
\input{table-heatmap-random.tex}
\end{table}

@@

\ref{fig:quantile-random}

@@

\begin{figure}[t]
  \caption{Quantile random.}
\label{fig:quantile-random}
\bigskip
\centering
\input{quantile-random.tex}
\end{figure}
